---
title: "Take-Home_Ex03"
format: html
---

## **Installing and Loading R packages**

```{r}
pacman::p_load(httr, tidyverse, sf, tmap, 
               jsonlite, progress, gtsummary, performance, RColorBrewer, sfdep, GWmodel)
```

## **Geocoding HDB resale data**

### **Selecting data of interest**

We will filter out 3 room, 4 room and 5 room flats

```{r}
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(flat_type %in% c("3 ROOM", "4 ROOM", "5 ROOM")) %>%
  filter(month >= "2025-09" & month < "2025-10")
```

We will create a function that helps us convert the address into latitude and longitude using OneMapSG api.

```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext<-content(res, as="text")
  
  output <- fromJSON(restext)  %>% 
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)

  return(output)
}
```

Now we will use that function

```{r}
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

```{r}
HDBresale %>% head()
```

We have the columns Latitude and Longitude.

## **Proximity Analysis**

Count number of geographic entities located within a defined distance of each HDB

Let's count the number of preschools located within 350m of each resale HDB

Convert hdb resale to sf

```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326) %>%
  st_transform(crs = 3414)
```

Import preschool data

```{r}
preschool = st_read(
  "data/geospatial/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)

CHASclinic = st_read(
  "data/geospatial/CHASClinics.geojson") %>%
  st_transform(crs= 3414)

childCareServices = st_read(
  "data/geospatial/ChildCareServices.geojson") %>%
  st_transform(crs= 3414)

hawkerCenters = st_read(
  "data/geospatial/HawkerCentres.geojson") %>%
  st_transform(crs= 3414)

parks = st_read(
  "data/geospatial/Parks.geojson") %>%
  st_transform(crs= 3414)

supermarkets = st_read(
  "data/geospatial/Supermarkets.geojson") %>%
  st_transform(crs= 3414)

eldercare <- st_read(
  dsn = "data/geospatial", 
  layer = "ELDERCARE"
) %>% 
  st_transform(crs = 3414)


#trainstation <- st_read(
 ##layer = "RapidTransitSystemStation"
#) %>% 
 #  st_make_valid() %>%   # convert multipolygon → point
#
  #st_transform(crs = 3414)


```

### **Counting number of preschools**

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_PRISCHOOL = map_int(within_350m, length))
```

### **Computing shortest distance**

```{r}
nearest_dist <- function(from_geom, to_sf) {
  as.numeric(
    st_distance(
      from_geom,
      to_sf[st_nearest_feature(from_geom, to_sf), ],
      by_element = TRUE
    )
  )
}

HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_PRESCHOOL     = nearest_dist(geometry, preschool),
    PROX_CHASCLINIC    = nearest_dist(geometry, CHASclinic),
    PROX_CHILDCARE     = nearest_dist(geometry, childCareServices),
    PROX_HAWKER        = nearest_dist(geometry, hawkerCenters),
    PROX_PARK          = nearest_dist(geometry, parks),
    PROX_SUPERMARKET   = nearest_dist(geometry, supermarkets),
    PROX_ELDERCARE     = nearest_dist(geometry, eldercare),
  )

```

```{r}
HDBresale_sf
```

## **13.5 Geospatial Data Wrangling**

### **13.5.1 Importing geospatial data**

```{r}
mpsz = st_read(dsn = "data/geospatial", 
               layer = "MP14_SUBZONE_WEB_PL")
```

### **13.5.2 Updating CRS information**

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

```{r}
st_crs(mpsz_svy21)
```

EPSG: is indicated as *3414* now.

```{r}
st_bbox(mpsz_svy21)
```

## **13.7 Exploratory Data Analysis (EDA)**

### **13.7.1 EDA using statistical graphics**

```{r}
ggplot(data=HDBresale_sf, 
       aes(x=`resale_price`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Right skewed meaning more hdb flats were transacted at lower relative prices.

Normalize the resale price using a new variable called *log_resale_price* by using a log transformation on the variable resale_price It is performed using `mutate()` of **dplyr** package.

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(`log_resale_price` = log(resale_price))
```

```{r}
ggplot(data=HDBresale_sf, 
       aes(x=`log_resale_price`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

distribution is relatively less skewed after the transformation.

### **13.7.2 Multiple Histogram Plots distribution of variables**

```{r}

library(purrr)

vars <- c("PROX_PRESCHOOL",
          "PROX_CHASCLINIC",
          "PROX_CHILDCARE",
          "PROX_HAWKER",
          "PROX_PARK",
          "PROX_SUPERMARKET",
          "PROX_ELDERCARE")
library(ggplot2)
library(purrr)
library(ggpubr)

plot_list <- map(vars, ~ {
  ggplot(HDBresale_sf, aes(x = .data[[.x]])) +
    geom_histogram(bins = 20, color = "black", fill = "lightblue") +
    labs(title = .x, x = .x, y = "Count")
})

ggarrange(plotlist = plot_list, 
          ncol = 3, 
          nrow = 3)


```

### **13.7.3 Drawing Statistical Point Map**

```{r}
tmap_mode("plot")
```

```{r}
tm_shape(mpsz_svy21)+
  tm_polygons() +
tm_shape(HDBresale_sf) +  
  tm_dots(fill = "resale_price",
          fill_alpha = 0.6,
          size = 0.5,
          fill.scale = tm_scale_intervals(
            style="quantile")) +
  tm_view(set_zoom_limits = c(11,14))
```

```{r}
tmap_mode("plot")
```

## **13.8 Hedonic Pricing Modelling in R**

building hedonic pricing models for HDB units using lm() of R base

### **13.8.1 Simple Linear Regression Method**

First, we will build a simple linear regression model by using *resale_price* as the dependent variable and *floor_area_sqm* as the independent variable.

```{r}
HDBslr <- lm(formula=resale_price ~ floor_area_sqm,
             data = HDBresale_sf)
```

`lm()` returns an object of class “lm” or for multiple responses of class c(“mlm”, “lm”).

The functions `summary()` and `anova()` can be used to obtain and print a summary and analysis of variance table of the results. The generic accessor functions coefficients, effects, fitted.values and residuals extract various useful features of the value returned by `lm`.

```{r}
summary(HDBslr)
```

The output report reveals that the SELLING_PRICE can be explained by using the formula:

```         
      *y = 122664.9 + 5701.3x*
```

The R-squared of 0.2987 reveals that the simple regression model built is able to explain about 29% of the resale prices.

Since p-value is much smaller than 0.0001, we will reject the null hypothesis that mean is a good estimator of resale_price. This will allow us to infer that simple linear regression model above is a good estimator of resale_price.

The **Coefficients:** section of the report reveals that the p-values of both the estimates of the Intercept and floor_area_sqm are smaller than 0.001. In view of this, the null hypothesis of the B0 and B1 are equal to 0 will be rejected. As a results, we will be able to infer that the B0 and B1 are good parameter estimates.

To visualise the best fit curve on a scatterplot, we can incorporate `lm()` as a method function in ggplot’s geometry as shown in the code chunk below.

```{r}
ggplot(data=HDBresale_sf,  
       aes(x=`floor_area_sqm`, y=`resale_price`)) +
  geom_point() +
  geom_smooth(method = lm)
```

Figure above reveals that there are a few statistical outliers with relatively high selling prices.

### **13.8.2 Multiple Linear Regression Method**

#### 13.8.2.1 Visualising the relationships of the independent variables

Before building a multiple regression model, it is important to ensure that the indepdent variables used are not highly correlated to each other. If these highly correlated independent variables are used in building a regression model by mistake, the quality of the model will be compromised. This phenomenon is known as **multicollinearity** in statistics.

The code chunk below is used to plot a scatterplot matrix of the relationship between the independent variables in *HDBresale* data.frame.

But before that we need to clean up some of our varaiables

Calculate age:

```{r}
library(stringr)
library(dplyr)

HDBresale_sf <- HDBresale_sf %>%
  mutate(
    # Extract years and months  
    lease_years  = as.numeric(str_extract(remaining_lease, "\\d+(?= years?)")),
    lease_months = as.numeric(str_extract(remaining_lease, "\\d+(?= months?)")),
    lease_months = ifelse(is.na(lease_months), 0, lease_months),
    
    # Convert to years
    remaining_lease_num = lease_years + lease_months / 12,
    
    # Compute AGE
    AGE = 99 - remaining_lease_num
  )

```

Convert flattype to numeric\

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    FLAT_TYPE_NUM = case_when(
      flat_type == "1 ROOM" ~ 1,
      flat_type == "2 ROOM" ~ 2,
      flat_type == "3 ROOM" ~ 3,
      flat_type == "4 ROOM" ~ 4,
      flat_type == "5 ROOM" ~ 5,
      flat_type == "EXECUTIVE" ~ 6,
      flat_type == "MULTI-GENERATION" ~ 7,
      TRUE ~ NA_real_
    )
  )

```

convert storey to numeric:

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    storey_low  = as.numeric(str_extract(storey_range, "^\\d+")),
    storey_high = as.numeric(str_extract(storey_range, "\\d+$")),
    STOREY_MID  = (storey_low + storey_high) / 2
  )

```

```{r}
library(corrplot)

num_vars <- HDBresale_sf %>%
  st_drop_geometry() %>%
  select(
    floor_area_sqm, AGE, FLAT_TYPE_NUM, STOREY_MID,
    starts_with("PROX"), 
    log_resale_price
  )

corr_matrix <- cor(num_vars, use = "complete.obs")

corrplot(corr_matrix,diag = FALSE, 
         order = "AOE",
         tl.pos = "td", 
         tl.cex = 0.5, 
         method = "number", 
         type = "upper")

```

### **13.8.3 Building a hedonic pricing model using multiple linear regression method**

```{r}
HDBmlr <- lm(
  formula = resale_price ~ floor_area_sqm + AGE  + 
    PROX_CHILDCARE + PROX_ELDERCARE +
    PROX_HAWKER + PROX_PARK +
    PROX_SUPERMARKET + 
    STOREY_MID + PROX_PRESCHOOL +
    FLAT_TYPE_NUM + PROX_CHASCLINIC,
  data=HDBresale_sf)
summary(HDBmlr)
```

### **13.8.4 Revising the model**

Let's remove the values that are not statistically significant.

PROX_CHILDCARE, PROX_PRESCHOOL, FLAT_TYPE_NUM since their p \< 0.05

```{r}
HDBmlr1 = lm(
  formula = resale_price ~ floor_area_sqm + AGE + PROX_ELDERCARE +
    PROX_HAWKER + PROX_PARK +
    PROX_SUPERMARKET + 
    STOREY_MID + PROX_CHASCLINIC,
  data=HDBresale_sf)

summary(HDBmlr1)
```

The output above reveals that all explanatory variables are statistically significant at 95% confident level.\

### **13.8.5 Preparing Publication Quality Table: gtsummary method**

```{r}
tbl_regression(HDBmlr1, intercept = TRUE)
```

```{r}
tbl_regression(HDBmlr1, 
               intercept = TRUE) %>% 
  add_glance_source_note(
    label = list(sigma ~ "\U03C3"),
    include = c(r.squared, adj.r.squared, 
                AIC, statistic,
                p.value, sigma))
```

### **13.8.6 Regression Diagnostics**

#### 13.8.6.1 Multicollinearity test

```{r}
check_collinearity(HDBmlr1)
```

```{r}
mlr.vif <- check_collinearity(HDBmlr1)
plot(mlr.vif)
```

Since the VIF of the independent variables are less than 10. We can safely conclude that there are no sign of multicollinearity among the independent variables.

#### 13.8.6.2 Test for Non-Linearity

```{r}
check_model(HDBmlr1,
            check = "linearity")
```

#### 13.8.6.3 Test for Normality Assumption

```{r}
check_normality(HDBmlr1)
```

The print above reveals that the p-value of the normality assumption test is less than alpha value of 0.05. Hence we reject the normality assumption at 95% confident level.

```{r}
plot(check_normality(HDBmlr1), 
     type = "qq")
```

Q-Q plot above below shows that majority of the data points are felt along the zero line.

```{r}
check_model(HDBmlr1,
            check= "normality")
```

The figure reveals that the residual of the multiple linear regression model (i.e. HDBmlr1) is resemble normal distribution

#### 13.8.6.4 Testing for Spatial Autocorrelation

```{r}
mlr.output <- as.data.frame(HDBmlr1$residuals)
```

```{r}
HDBresale_sf <- cbind(HDBresale_sf, 
                        HDBmlr1$residuals) %>%
rename(`MLR_RES` = `HDBmlr1.residuals`)
```

```{r}
tmap_mode("plot")
```

```{r}
tm_shape(mpsz_svy21) +
  tm_polygons(fill_alpha = 0.4) +
tm_shape(HDBresale_sf) +  
  tm_dots(
    fill = "MLR_RES",
    size = 0.7,
    col = "black",
    fill.scale = tm_scale(
      n = 10,
      values = rev(
        brewer.pal(11, "RdBu")),
      style = "quantile",
      midpoint = NA),
    fill.legend = tm_legend(title = "Residuals")
  ) + 
  tm_title("LM Residuals (Quantile Classification)") +
  tm_layout(legend.outside = TRUE) +
  tm_view(set_zoom_limits = c(11,14))
```

```{r}
tmap_mode("plot")
```

The figure above reveal that there is sign of spatial autocorrelation.

To proof that our observation is indeed true, Global Moran’s I test will be performed

First, we will compute the distance-based weight matrix by using [`st_dist_band()`](https://sfdep.josiahparry.com/reference/st_dist_band) function of **sfdep**.

```{r}
HDBresale_sf <- HDBresale_sf %>% 
  mutate(
    nb = st_dist_band(
      st_geometry(geometry), 
                 upper = 1500),
    wt = st_weights(
      nb, style = "W"),
    .before = 1) 
```

```{r}
set.seed(1234)
global_moran_perm(
  HDBresale_sf$MLR_RES, 
  nb = HDBresale_sf$nb, 
  wt = HDBresale_sf$wt,
  alternative = "two.sided", 
  nsim = 499)
```

The Global Moran’s I test for residual spatial autocorrelation shows that it’s p-value is less than 0.00000000000000022 which is less than the alpha value of 0.05. Hence, we will reject the null hypothesis that the residuals are randomly distributed.

Since the Observed Global Moran I = 0.14389 which is greater than 0, we can infer than the residuals resemble cluster distribution.

## **13.9 Building Hedonic Pricing Models using GWmodel**

### **13.9.1 Building Fixed Bandwidth GWR Model**

#### 13.9.1.1 Computing fixed bandwith

```{r}
bw.fixed <- bw.gwr(
  formula = resale_price ~ floor_area_sqm + AGE + PROX_ELDERCARE +
    PROX_HAWKER + PROX_PARK +
    PROX_SUPERMARKET + 
    STOREY_MID + PROX_CHASCLINIC,
  data=HDBresale_sf, 
  approach="CV", 
  kernel="gaussian", 
  adaptive=FALSE, 
  longlat=FALSE)
```

#### 13.9.1.2 GWModel method - fixed bandwith

```{r}
gwr.fixed <- gwr.basic(
  formula = resale_price ~ floor_area_sqm + AGE + PROX_ELDERCARE +
    PROX_HAWKER + PROX_PARK +
    PROX_SUPERMARKET + 
    STOREY_MID + PROX_CHASCLINIC,
  data=HDBresale_sf, 
  bw=bw.fixed, 
  kernel = 'gaussian', 
  longlat = FALSE)
```

```{r}
gwr.fixed
```

The report shows that the AICc of the gwr is 48142.36 which is significantly smaller than the globel multiple linear regression model of 51300.11

### **13.9.2 Building Adaptive Bandwidth GWR Model**

#### 13.9.2.1 Computing the adaptive bandwidth

```{r}
bw.adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm + AGE + PROX_ELDERCARE +
    PROX_HAWKER + PROX_PARK +
    PROX_SUPERMARKET + 
    STOREY_MID + PROX_CHASCLINIC,
  data=HDBresale_sf,
  approach="CV", 
  kernel="gaussian", 
  adaptive=TRUE, 
  longlat=FALSE)
```

#### 13.9.2.2 Constructing the adaptive bandwidth gwr model

```{r}
gwr.adaptive <- gwr.basic(
  formula = resale_price ~ floor_area_sqm + AGE + PROX_ELDERCARE +
    PROX_HAWKER + PROX_PARK +
    PROX_SUPERMARKET + 
    STOREY_MID + PROX_CHASCLINIC,
  data=HDBresale_sf,
  bw=bw.adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE, 
  longlat = FALSE)
```

```{r}
gwr.adaptive
```

The report shows that the AICc the adaptive distance gwr is 48440.2 which is even smaller than the AICc of the fixed distance gwr of 48142.36 .

### **13.9.3 Visualising GWR Output**

### **13.9.4 Converting SDF into *sf* data.frame**

```{r}
HDBresale_sf.adaptive <-
  st_as_sf(gwr.adaptive$SDF) %>%
  st_transform(crs=3414)
```

```{r}
glimpse(HDBresale_sf.adaptive)
```

```{r}
summary(gwr.adaptive$SDF$yhat)
```

### **13.9.5 Visualising local R2**

```{r}
tmap_mode("plot")
tm_shape(mpsz_svy21)+
  tm_polygons(fill_alpha = 0.1) +
tm_shape(HDBresale_sf.adaptive) +  
  tm_dots(col = "Local_R2",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

tmap_mode("plot")
```

### **13.9.6 Visualising coefficient estimates**

```{r}
tmap_mode("plot")
AREA_SQM_SE <- tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(HDBresale_sf.adaptive) +  
  tm_dots(col = "floor_area_sqm_SE",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

AREA_SQM_TV <- tm_shape(mpsz_svy21)+
  tm_polygons(alpha = 0.1) +
tm_shape(HDBresale_sf.adaptive) +  
  tm_dots(col = "floor_area_sqm_TV",
          border.col = "gray60",
          border.lwd = 1) +
  tm_view(set.zoom.limits = c(11,14))

tmap_arrange(AREA_SQM_SE, AREA_SQM_TV, 
             asp=1, ncol=2,
             sync = TRUE)

tmap_mode("plot")
```
