---
title: "Hands-on Excercise 3"
date: "5/9/25"
date-modified: "last-modified"
format: html
execute: 
  echo: true
  eval: true
  warning: false
  freeze: true
---

```{r}
pacman::p_load(sf, terra, spatstat, sparr, tmap, tidyverse)
```

## **6.5 Importing and Preparing Study Area**

### **6.5.1 Importing study area**

```{r}
kbb <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") 
```

```{r}
kbb_sf <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

### **6.5.2 Converting OWIN**

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

```{r}
class(kbb_owin)
```

## **6.6 Importing and Preparing Forest Fire data**

Next, we will import the forest fire data set (i.e. *forestfires.csv*) into R environment.

```{r}
fire_sf <- read_csv("data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
                       crs = 4326) %>%
  st_transform(crs = 32748)
```

convert data type of *acq_date* to numeric.

```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

## **6.7 Visualising the Fire Points**

### **6.7.1 Overall plot**

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

### **6.7.2 Visuaising geographic distribution of forest fires by month**

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

## **6.8 Computing STKDE by Month**

### **6.8.1 Extracting forest fires by month**

```{r}
fire_month <- fire_sf %>% 
  select(Month_num)
```

### **6.8.2 Creating ppp**

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

```{r}
summary(fire_month_ppp)
```

check for duplicates

```{r}
any(duplicated(fire_month_ppp))
```

### **6.8.3 Including Owin object**

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

```{r}
plot(fire_month_owin)
```

### **6.8.4 Computing Spatio-temporal KDE**

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

### **6.8.5 Plotting the spatio-temporal KDE object**

```{r}
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

## **6.9 Computing STKDE by Day of Year**

### **6.9.1 Creating ppp object**

```{r}
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

### **6.9.2 Including Owin object**

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

### **6.9.3**

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```

```{r}
plot(kde_yday)

```

## **6.10 Computing STKDE by Day of Year: Improved method**

One of the nice function provides in sparr package is `BOOT.spattemp()`. It support bandwidth selection for standalone spatiotemporal density/intensity based on bootstrap estimation of the MISE, providing an isotropic scalar spatial bandwidth and a scalar temporal bandwidth.

Code chunk below uses `BOOT.spattemp()` to determine both the spatial bandwidth and the scalar temporal bandwidth.

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin) 
```

### **6.10.1 Computing spatio-temporal KDE**

```{r}
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 9000,
                             lambda = 19)
summary(kde_yday)
```

### **6.10.2 Plotting the output spatio-temporal KDE**

```{r}
plot(kde_yday)
```
