---
title: "extractingRegions"
format: html
---

```{r}
pacman::p_load(tidyverse, sf, tmap, httr, spatstat, rvest, sparr, stpp)
```

```{r}
biz47 <- readRDS("data/rds/biz_47.rds")
biz551 <- readRDS("data/rds/biz_551.rds")
biz932 <- readRDS("data/rds/biz_932.rds")
```

Convert tibble to sf

```{r}

biz47_sf <- st_as_sf(biz47,
                     coords = c("X", "Y"),
                     crs = 3414)

biz551_sf <- st_as_sf(biz551,
                      coords = c("X", "Y"),
                      crs = 3414)

biz932_sf <- st_as_sf(biz932,
                      coords = c("X", "Y"),
                      crs = 3414)
```

Extract regions from mpsz

```{r} 
extract_kml_field <- function(html_text, field_name) { 
  if (is.na(html_text) || html_text == "") 
    return(NA_character_)
  
  page <- read_html(html_text) 
  rows <- page %>% html_elements("tr")
  
  value <- rows %>% 
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>% 
    html_element("td") %>% 
    html_text2()
  
  if (length(value) == 0) NA_character_ else value 
}
```


```{r}
mpsz_sf <- st_read("data/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)
```


```{r} 
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N   = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N  = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C  = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

We parse out REGION, PLANNING AREA, and SUBZONE attributes.

```{r} 
# remove islands
mpsz_cl <- mpsz_sf %>%
  filter(
    SUBZONE_N != "SOUTHERN GROUP",
    PLN_AREA_N != "WESTERN ISLANDS",
    PLN_AREA_N != "NORTH-EASTERN ISLANDS"
  )
```
Save cleaned dataset

Write the filtered shapefile to RDS for later use.
```{r}
write_rds(mpsz_cl, "data/rds/mpsz_cl.rds")
```

Visualization
Retail Trade (47)
```{r}
# --- Retail Trade (47) --- 
tm_shape(mpsz_cl) + tm_polygons() + tm_shape(biz47_sf) + tm_dots(col = "red")
```
Accommodation (551)
```{r}
# --- Accommodation (551) --- 
tm_shape(mpsz_cl) + tm_polygons() + tm_shape(biz551_sf) + tm_dots(col = "blue")
```
Amusement & Recreation (932)
```{r}
# --- Amusement & Recreation (932) --- 
tm_shape(mpsz_cl) + tm_polygons() + tm_shape(biz932_sf) + tm_dots(col = "green")
```
Since majority of the points are concentrated in the Central Region, we can zoom into that area for better visualization.

```{r}
# Zoom into Central Region
# filter central region
centralarea <- mpsz_cl %>%
  filter(REGION_N == "CENTRAL REGION")

# make sure CRS matches
biz47_sf <- st_transform(biz47_sf, st_crs(centralarea))
```

Retail Trade (47)
```{r}
# --- Retail Trade (47) ---

biz47_sf_in <- st_intersection(biz47_sf, centralarea)
tm_shape(centralarea) + tm_polygons() + tm_shape(biz47_sf_in) + tm_dots(col = "red")
```

Accommodation (551)
```{r}
# --- Accommodation (551) ---

# make sure CRS matches
biz551_sf <- st_transform(biz551_sf, st_crs(centralarea))

biz551_sf_in <- st_intersection(biz551_sf, centralarea)
tm_shape(centralarea) + tm_polygons() + tm_shape(biz551_sf_in) + tm_dots(col = "blue")
```
Amusement & Recreation (932)
```{r}
# --- Amusement & Recreation (932) ---

# make sure CRS matches
biz932_sf <- st_transform(biz932_sf, st_crs(centralarea))

biz932_sf_in <- st_intersection(biz932_sf, centralarea)

tm_shape(centralarea) + tm_polygons() + tm_shape(biz932_sf_in) + tm_dots(col = "green")
```

From the visualizations, we can see that the points are concentrated in the Central Region, particularly in areas such as Orchard, Marina Bay, and Bugis. This indicates that these areas are popular for retail trade, accommodation, and amusement & recreation activities. The spatial distribution of these business activities can provide insights into consumer behavior and preferences in Singapore.


perform KDE on the central region
```{r}
# Convert to sf object

centralarea <- as.owin(centralarea)

biz47_ppp  <- as.ppp(st_coordinates(biz47_sf_in),  W = centralarea)
biz551_ppp <- as.ppp(st_coordinates(biz551_sf_in), W = centralarea)
biz932_ppp <- as.ppp(st_coordinates(biz932_sf_in), W = centralarea)
```
### Kernel density across Singapore for all businesses

```{r}
# --- biz47 ---
bw_diggle_47 <- bw.diggle(biz47_ppp)
kde_biz47 <- density(
  biz47_ppp,
  sigma = bw_diggle_47,
  edge = TRUE,
  kernel = "gaussian"
)

# --- biz551 ---
bw_diggle_551 <- bw.diggle(biz551_ppp)
kde_biz551 <- density(
  biz551_ppp,
  sigma = bw_diggle_551,
  edge = TRUE,
  kernel = "gaussian"
)

# --- biz932 ---
bw_diggle_932 <- bw.diggle(biz932_ppp)
kde_biz932 <- density(
  biz932_ppp,
  sigma = bw_diggle_932,
  edge = TRUE,
  kernel = "gaussian"
)

```

### KDE for Retail (47)

```{r}
kde_biz47 <- density(
  biz47_ppp,
  sigma=1000, 
  edge=TRUE,
  kernel="gaussian"
)
plot(kde_biz47)
```

### KDE for Accommodation (551)

```{r}
kde_biz551 <- density(
  biz551_ppp,
  sigma = 1000, 
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_biz551, main = "KDE - Accommodation (551)")
```

### KDE - Amusement & Recreation (932)

```{r}
kde_biz932 <- density(
  biz932_ppp,
  sigma = 1000, 
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_biz932, main = "KDE - Amusement & Recreation (932)")
```
## Spatio Temporal KDE by months
We check how many businesses were opened each month and their locaitons.

### Retail Trade (47)

```{r}


# 1) Build ordered month index (numeric) + nice label
biz47_sf_in <- biz47_sf_in %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),
    YearMonthLab  = format(YearMonthDate, "%Y-%m"),
    Month_idx     = as.integer(factor(YearMonthDate,
                                      levels = sort(unique(YearMonthDate))))
  )

# 2) Keep only Month_idx + geometry
biz47_month <- biz47_sf_in %>%
  select(Month_idx)

# 3) Convert to ppp
biz47_month_ppp <- as.ppp(biz47_month)
summary(biz47_month_ppp)
any(duplicated(biz47_month_ppp))

# 4) Clip to Central Region only
central_owin <- as.owin(centralarea)
biz47_month_central <- biz47_month_ppp[central_owin]
summary(biz47_month_central)

# 5) Plot
plot(biz47_month_central)
```
```{r}
biz47_stkde <- spattemp.density(biz47_month_central)
summary(biz47_stkde)
```
```{r}
# 6) Plot all months sequentially, full size, with proper YYYY-MM labels
#    Build a lookup from index -> label
idx_levels <- sort(unique(biz47_sf_in$Month_idx))
lab_lookup <- biz47_sf_in %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)


for (i in idx_levels) {
  plot(biz47_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz47 (", lab_lookup$YearMonthLab[lab_lookup$Month_idx == i], ")", sep = ""))
}


```


### Short-Term Accomodation (551)

```{r}
# 1) Build ordered month index (numeric) + nice label
biz551_sf_in <- biz551_sf_in %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),
    YearMonthLab  = format(YearMonthDate, "%Y-%m"),
    Month_idx     = as.integer(factor(YearMonthDate,
                                      levels = sort(unique(YearMonthDate))))
  )
# 2) Keep only Month_idx + geometry
biz551_month <- biz551_sf_in %>%
  select(Month_idx)
# 3) Convert to ppp
biz551_month_ppp <- as.ppp(biz551_month)
summary(biz551_month_ppp)
any(duplicated(biz551_month_ppp))
# 4) Clip to Central Region only
central_owin <- as.owin(centralarea)
biz551_month_central <- biz551_month_ppp[central_owin]
summary(biz551_month_central)
# 5) Plot
plot(biz551_month_central)
``` 

```{r}
biz551_stkde <- spattemp.density(biz551_month_central)
summary(biz551_stkde)
```
```{r}
# 6) Plot all months sequentially, full size, with proper YYYY-MM labels
#    Build a lookup from index -> label
  
idx_levels <- sort(unique(biz551_sf_in$Month_idx))
lab_lookup <- biz551_sf_in %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)
for (i in idx_levels) {
    
  plot(biz551_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz551 (", lab_lookup$YearMonthLab[lab_lookup$Month_idx == i], ")", sep = ""))
}
```
### Amusement & Recreation (932)
```{r}
# 1) Build ordered month index (numeric) + nice label
biz932_sf_in <- biz932_sf_in %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),
    YearMonthLab  = format(YearMonthDate, "%Y-%m"),
    Month_idx     = as.integer(factor(YearMonthDate,
                                      levels = sort(unique(YearMonthDate))))
  )

# 2) Keep only Month_idx + geometry
biz932_month <- biz932_sf_in %>%
  select(Month_idx)

# 3) Convert to ppp
biz932_month_ppp <- as.ppp(biz932_month)
summary(biz932_month_ppp)
any(duplicated(biz932_month_ppp))

# 4) Clip to Central Region only
central_owin <- as.owin(centralarea)
biz932_month_central <- biz932_month_ppp[central_owin]
summary(biz932_month_central)

# 5) Plot
plot(biz932_month_central)
```

```{r}
# Spatio-temporal KDE for attractions
biz932_stkde <- spattemp.density(biz932_month_central)

# Summarise results
summary(biz932_stkde)
```
```{r}
# 6) Plot all months sequentially, full size, with proper YYYY-MM labels
#    Build a lookup from index -> label

idx_levels <- sort(unique(biz932_sf_in$Month_idx))
lab_lookup <- biz932_sf_in %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)

for (i in idx_levels) {
  plot(biz932_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz932 (", 
                    lab_lookup$YearMonthLab[lab_lookup$Month_idx == i], 
                    ")", sep = ""))
}
```
From the spatio-temporal KDE plots, we can observe the following trends for each business category over time:\
1. **Retail Trade (47)**: The density of retail trade businesses appears to increase steadily over time, with notable concentrations in areas such as Orchard and Marina Bay. This suggests a growing demand for retail services in these popular shopping districts.
2. **Accommodation (551)**: The density of accommodation businesses shows a more fluctuating pattern, with peaks during certain months. This could be indicative of seasonal tourism trends, where certain times of the year see higher demand for accommodation services, particularly in areas like Marina Bay and Bugis.
3. **Amusement & Recreation (932)**: The density of amusement and recreation businesses also shows variability over time, with certain months exhibiting higher concentrations. This may reflect the opening of new attractions or seasonal events that draw visitors to specific areas.
Overall, the spatio-temporal KDE analysis provides valuable insights into the dynamics of business activities in Singapore, highlighting areas of growth and seasonal trends that can inform business strategies and urban planning efforts.

## 2nd order spatial Analysis

