---
title: "Take-Home_Ex01"
author: "Mohammad Farhan Ahsan"
date: "September 20, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---


```{r}
#Importing Libraries

pacman::p_load(tidyverse, sf, tmap, httr, spatstat, rvest, sparr, stpp)
```


# 1. Datasets used

 - Acra data (Selected business types: Retail Trade (47), Short Term Accomodation (551), Other Amusement and Recreation Activities)

 - Master Plan 2019 Subzone Boundary (No Sea), a polygon feature data providing information of URA 2019 Master Plan Planning Subzone boundary data.
 
I have selected the following Business types. Studying geospatial data and analysing these businesses will help us get a clear picture of how Singapore's topurism scene functions. These analysis can be used for new businesses in future to position themselves intelligently to increase revenue. It can also be used for urban planning.

# 2. Data Cleaning

```{r, eval=FALSE}
folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path, 
                        pattern = "^ACRA*.*\\.csv$", 
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)
```

pad problematc postal codes
```{r, eval=FALSE}
acra_data <- acra_data %>% mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0"))
```

We will keep the first 25 columns and filter for businesses classified under SSIC codes **47 (Retail Trade)**, **551 (Short Term Accommodation Activities)**, and **932 (Other Amusement and Recreation Activities)**. Additionally, we will restrict the dataset to businesses registered between **1 January 2024** and **30 June 2025**.

```{r, eval=FALSE}

# Retail Trade (47)
biz47 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 2) == "47"
  )

# Short Term Accommodation Activities (551)
biz551 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 3) == "551"
  )

# Other Amusement and Recreation Activities (932)
biz932 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 3) == "932"
  )

```

## Fetch all unique postal

```{r, eval=FALSE}
# 551 - Short Term Accommodation Activities
postcodes_551 <- unique(biz551$postal_code)

# 47 - Retail Trade
postcodes_47 <- unique(biz47$postal_code)

# 932 - Other Amusement and Recreation Activities
postcodes_932 <- unique(biz932$postal_code)

```
## Geocoding

```{r, eval=FALSE}
library(httr)
library(dplyr)

fetch_postcodes <- function(df, url) {
  postcodes <- unique(df$postal_code)
  
  found <- data.frame()
  not_found <- data.frame(postcode = character())
  
  for (pc in postcodes) {
    query <- list(
      searchVal = pc,
      returnGeom = "Y",
      getAddrDetails = "Y",
      pageNum = "1"
    )
    
    res <- GET(url, query = query)
    json <- content(res)
    
    if (json$found != 0) {
      df_res <- as.data.frame(json$results, stringsAsFactors = FALSE)
      df_res$input_postcode <- pc
      found <- bind_rows(found, df_res)
    } else {
      not_found <- bind_rows(not_found, data.frame(postcode = pc))
    }
  }
  
  return(list(found = found, not_found = not_found))
}

```

```{r, eval=FALSE}
url <- "https://onemap.gov.sg/api/common/elastic/search"

res_47  <- fetch_postcodes(biz47, url)
res_551 <- fetch_postcodes(biz551, url)
res_932 <- fetch_postcodes(biz932, url)

```

Fetch the first 10 columns including the lat and long

```{r, eval=FALSE}
res_47$found <- res_47$found %>% select(1:10)
res_551$found <- res_551$found %>% select(1:10)
res_932$found <- res_932$found %>% select(1:10)
```

Appending the location information

```{r, eval=FALSE}
biz47 <- biz47 %>%
  left_join(res_47$found, 
            by = c("postal_code" = "POSTAL"))

biz551 <- biz551 %>%
  left_join(res_551$found, 
            by = c("postal_code" = "POSTAL"))

biz932 <- biz932 %>%
  left_join(res_932$found, 
            by = c("postal_code" = "POSTAL"))

```

Save the data

```{r, eval=FALSE}
write_rds(biz47, "data/rds/biz_47.rds")
write_rds(biz551, "data/rds/biz_551.rds")
write_rds(biz932, "data/rds/biz_932.rds")

```

# READING
```{r}
biz47 <- readRDS("data/rds/biz_47.rds")
biz551 <- readRDS("data/rds/biz_551.rds")
biz932 <- readRDS("data/rds/biz_932.rds")
```

Convert tibble to sf

```{r}

biz47_sf <- st_as_sf(biz47,
                     coords = c("X", "Y"),
                     crs = 3414)

biz551_sf <- st_as_sf(biz551,
                      coords = c("X", "Y"),
                      crs = 3414)

biz932_sf <- st_as_sf(biz932,
                      coords = c("X", "Y"),
                      crs = 3414)
```

Extract regions from mpsz

```{r}
extract_kml_field <- function(html_text, field_name) { 
  if (is.na(html_text) || html_text == "") 
    return(NA_character_)
  
  page <- read_html(html_text) 
  rows <- page %>% html_elements("tr")
  
  value <- rows %>% 
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>% 
    html_element("td") %>% 
    html_text2()
  
  if (length(value) == 0) NA_character_ else value 
}
```

```{r}
mpsz_sf <- st_read("data/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)
```

```{r}
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N   = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N  = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C  = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

We parse out REGION, PLANNING AREA, and SUBZONE attributes.

```{r}
# remove islands
mpsz_cl <- mpsz_sf %>%
  filter(
    SUBZONE_N != "SOUTHERN GROUP",
    PLN_AREA_N != "WESTERN ISLANDS",
    PLN_AREA_N != "NORTH-EASTERN ISLANDS"
  )
```

Save cleaned dataset

Write the filtered shapefile to RDS for later use.

```{r}
write_rds(mpsz_cl, "data/rds/mpsz_cl.rds")
```

Visualization Retail Trade (47)

```{r}
# --- Retail Trade (47) --- 
tm_shape(mpsz_cl) + tm_polygons() + tm_shape(biz47_sf) + tm_dots(col = "red")
```

Accommodation (551)

```{r}
# --- Accommodation (551) --- 
tm_shape(mpsz_cl) + tm_polygons() + tm_shape(biz551_sf) + tm_dots(col = "blue")
```

Amusement & Recreation (932)

```{r}
# --- Amusement & Recreation (932) --- 
tm_shape(mpsz_cl) + tm_polygons() + tm_shape(biz932_sf) + tm_dots(col = "green")
```

Since majority of the points are concentrated in the Central Region, we can zoom into that area for better visualization.

```{r}
# Zoom into Central Region
# filter central region
centralarea <- mpsz_cl %>%
  filter(REGION_N == "CENTRAL REGION")

# make sure CRS matches
biz47_sf <- st_transform(biz47_sf, st_crs(centralarea))
```

Retail Trade (47)

```{r}
# --- Retail Trade (47) ---

biz47_sf_in <- st_intersection(biz47_sf, centralarea)
tm_shape(centralarea) + tm_polygons() + tm_shape(biz47_sf_in) + tm_dots(col = "red")
```

Accommodation (551)

```{r}
# --- Accommodation (551) ---

# make sure CRS matches
biz551_sf <- st_transform(biz551_sf, st_crs(centralarea))

biz551_sf_in <- st_intersection(biz551_sf, centralarea)
tm_shape(centralarea) + tm_polygons() + tm_shape(biz551_sf_in) + tm_dots(col = "blue")
```

Amusement & Recreation (932)

```{r}
# --- Amusement & Recreation (932) ---

# make sure CRS matches
biz932_sf <- st_transform(biz932_sf, st_crs(centralarea))

biz932_sf_in <- st_intersection(biz932_sf, centralarea)

tm_shape(centralarea) + tm_polygons() + tm_shape(biz932_sf_in) + tm_dots(col = "green")
```

From the visualizations, we can see that the points are concentrated in the Central Region, particularly in areas such as Orchard, Marina Bay, and Bugis. This indicates that these areas are popular for retail trade, accommodation, and amusement & recreation activities. The spatial distribution of these business activities can provide insights into consumer behavior and preferences in Singapore.

perform KDE on the central region

```{r}
# Convert to sf object

centralarea <- as.owin(centralarea)

biz47_ppp  <- as.ppp(st_coordinates(biz47_sf_in),  W = centralarea)
biz551_ppp <- as.ppp(st_coordinates(biz551_sf_in), W = centralarea)
biz932_ppp <- as.ppp(st_coordinates(biz932_sf_in), W = centralarea)
```

### Kernel density across Singapore for all businesses

```{r}
# --- biz47 ---
bw_diggle_47 <- bw.diggle(biz47_ppp)
kde_biz47 <- density(
  biz47_ppp,
  sigma = bw_diggle_47,
  edge = TRUE,
  kernel = "gaussian"
)

# --- biz551 ---
bw_diggle_551 <- bw.diggle(biz551_ppp)
kde_biz551 <- density(
  biz551_ppp,
  sigma = bw_diggle_551,
  edge = TRUE,
  kernel = "gaussian"
)

# --- biz932 ---
bw_diggle_932 <- bw.diggle(biz932_ppp)
kde_biz932 <- density(
  biz932_ppp,
  sigma = bw_diggle_932,
  edge = TRUE,
  kernel = "gaussian"
)

```

### KDE for Retail (47)

```{r}
kde_biz47 <- density(
  biz47_ppp,
  sigma=1000, 
  edge=TRUE,
  kernel="gaussian"
)
plot(kde_biz47)
```

### KDE for Accommodation (551)

```{r}
kde_biz551 <- density(
  biz551_ppp,
  sigma = 1000, 
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_biz551, main = "KDE - Accommodation (551)")
```

### KDE - Amusement & Recreation (932)

```{r}
kde_biz932 <- density(
  biz932_ppp,
  sigma = 1000, 
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_biz932, main = "KDE - Amusement & Recreation (932)")
```

## Spatio Temporal KDE by months

We check how many businesses were opened each month and their locaitons.

### Retail Trade (47)

```{r}


# 1) Build ordered month index (numeric) + nice label
biz47_sf_in <- biz47_sf_in %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),
    YearMonthLab  = format(YearMonthDate, "%Y-%m"),
    Month_idx     = as.integer(factor(YearMonthDate,
                                      levels = sort(unique(YearMonthDate))))
  )

# 2) Keep only Month_idx + geometry
biz47_month <- biz47_sf_in %>%
  select(Month_idx)

# 3) Convert to ppp
biz47_month_ppp <- as.ppp(biz47_month)
summary(biz47_month_ppp)
any(duplicated(biz47_month_ppp))

# 4) Clip to Central Region only
central_owin <- as.owin(centralarea)
biz47_month_central <- biz47_month_ppp[central_owin]
summary(biz47_month_central)

# 5) Plot
plot(biz47_month_central)
```

```{r}
biz47_stkde <- spattemp.density(biz47_month_central)
summary(biz47_stkde)
```

```{r}
# 6) Plot all months sequentially, full size, with proper YYYY-MM labels
#    Build a lookup from index -> label
idx_levels <- sort(unique(biz47_sf_in$Month_idx))
lab_lookup <- biz47_sf_in %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)


for (i in idx_levels) {
  plot(biz47_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz47 (", lab_lookup$YearMonthLab[lab_lookup$Month_idx == i], ")", sep = ""))
}


```

### Short-Term Accomodation (551)

```{r}
# 1) Build ordered month index (numeric) + nice label
biz551_sf_in <- biz551_sf_in %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),
    YearMonthLab  = format(YearMonthDate, "%Y-%m"),
    Month_idx     = as.integer(factor(YearMonthDate,
                                      levels = sort(unique(YearMonthDate))))
  )
# 2) Keep only Month_idx + geometry
biz551_month <- biz551_sf_in %>%
  select(Month_idx)
# 3) Convert to ppp
biz551_month_ppp <- as.ppp(biz551_month)
summary(biz551_month_ppp)
any(duplicated(biz551_month_ppp))
# 4) Clip to Central Region only
central_owin <- as.owin(centralarea)
biz551_month_central <- biz551_month_ppp[central_owin]
summary(biz551_month_central)
# 5) Plot
plot(biz551_month_central)
```

```{r}
biz551_stkde <- spattemp.density(biz551_month_central)
summary(biz551_stkde)
```

```{r}
# 6) Plot all months sequentially, full size, with proper YYYY-MM labels
#    Build a lookup from index -> label
  
idx_levels <- sort(unique(biz551_sf_in$Month_idx))
lab_lookup <- biz551_sf_in %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)
for (i in idx_levels) {
    
  plot(biz551_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz551 (", lab_lookup$YearMonthLab[lab_lookup$Month_idx == i], ")", sep = ""))
}
```

### Amusement & Recreation (932)

```{r}
# 1) Build ordered month index (numeric) + nice label
biz932_sf_in <- biz932_sf_in %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),
    YearMonthLab  = format(YearMonthDate, "%Y-%m"),
    Month_idx     = as.integer(factor(YearMonthDate,
                                      levels = sort(unique(YearMonthDate))))
  )

# 2) Keep only Month_idx + geometry
biz932_month <- biz932_sf_in %>%
  select(Month_idx)

# 3) Convert to ppp
biz932_month_ppp <- as.ppp(biz932_month)
summary(biz932_month_ppp)
any(duplicated(biz932_month_ppp))

# 4) Clip to Central Region only
central_owin <- as.owin(centralarea)
biz932_month_central <- biz932_month_ppp[central_owin]
summary(biz932_month_central)

# 5) Plot
plot(biz932_month_central)
```

```{r}
# Spatio-temporal KDE for attractions
biz932_stkde <- spattemp.density(biz932_month_central)

# Summarise results
summary(biz932_stkde)
```

```{r}
# 6) Plot all months sequentially, full size, with proper YYYY-MM labels
#    Build a lookup from index -> label

idx_levels <- sort(unique(biz932_sf_in$Month_idx))
lab_lookup <- biz932_sf_in %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)

for (i in idx_levels) {
  plot(biz932_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz932 (", 
                    lab_lookup$YearMonthLab[lab_lookup$Month_idx == i], 
                    ")", sep = ""))
}
```

From the spatio-temporal KDE plots, we can observe the following trends for each business category over time:\
1. **Retail Trade (47)**: The density of retail trade businesses appears to increase steadily over time, with notable concentrations in areas such as Orchard and Marina Bay. This suggests a growing demand for retail services in these popular shopping districts. 2. **Accommodation (551)**: The density of accommodation businesses shows a more fluctuating pattern, with peaks during certain months. This could be indicative of seasonal tourism trends, where certain times of the year see higher demand for accommodation services, particularly in areas like Marina Bay and Bugis. 3. **Amusement & Recreation (932)**: The density of amusement and recreation businesses also shows variability over time, with certain months exhibiting higher concentrations. This may reflect the opening of new attractions or seasonal events that draw visitors to specific areas. Overall, the spatio-temporal KDE analysis provides valuable insights into the dynamics of business activities in Singapore, highlighting areas of growth and seasonal trends that can inform business strategies and urban planning efforts.

## 2nd order spatial Analysis

Using F function for retail businesses (47)

```{r, eval=FALSE}
F_biz47 = Fest(biz47_ppp)
plot(F_biz47, main = "F function - Retail Trade (47)")
```

Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

Ho = The distribution of retail shops in CBD are randomly distributed.

H1= The distribution of retail shops in CBD are not randomly distributed.

The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

```{r, eval=FALSE}
envelope_F_biz47 = envelope(biz47_ppp, Fest, nsim=99)
```

```{r, eval=FALSE}
plot(envelope_F_biz47, main = "F function - Retail Trade (47) with CSR envelope")
```

### Using K function for Short term Accomodations (551)
We chose K funcion so that we can see how closely packed the short term accommodations are in the CBD area.

```{r, eval=FALSE}
K_biz551 = Kest(biz551_ppp, correction = "Ripley")
plot(K_biz551, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```
This indicates clustering 


### Performing Complete Spatial Randomness Test
To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypotheses are as follows:

H₀: The distribution of short-term accommodations in the CBD is randomly distributed (follows CSR).

H₁: The distribution of short-term accommodations in the CBD is not randomly distributed (deviates from CSR).

The null hypothesis will be rejected if the p-value is smaller than the alpha value of 0.001.

```{r, eval=FALSE}
envelope_K_biz551 <- envelope(biz551_ppp, Kest, nsim=99, rank = 1, glocal=TRUE)
```

```{r, eval=FALSE}

plot(envelope_K_biz551, main = "K function - Short Term Accomodation (551) with CSR envelope")

```
### L function for Amusement & Recreation (932)
We chose K funcion so that we can see how closely packed the amusement & recreation businesses are in the CBD area.
```{r, eval=FALSE}
L_biz932 = Lest(biz932_ppp, correction = "Ripley")
plot(L_biz932, . -r ~ r, ylab= "L(d)-r", xlab = "d(m)")
```
#### Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypotheses are as follows:

H₀: The distribution of attractions in the central area is randomly distributed (follows CSR).

H₁: The distribution of attractions in the central area is not randomly distributed (deviates from CSR).

The null hypothesis will be rejected if the p-value is smaller than the alpha value of 0.001.

```{r, eval=FALSE}
envelope_L_biz932 <- envelope(biz932_ppp, Lest, nsim=99, rank = 1, glocal=TRUE)
```

```{r, eval=FALSE}
plot(envelope_L_biz932, main = "L function - Amusement & Recreation (932) with CSR envelope")
```



