---
title: "Take-Home_Ex01"
author: "Mohammad Farhan Ahsan"
date: "September 20, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Datasets used

Acra data

*Master Plan 2019 Subzone Boundary (No Sea)*, a polygon feature data providing information of URA 2019 Master Plan Planning Subzone boundary data.

## Business types selected:

47: Retail Trade

55: SHORT TERM ACCOMMODATION ACTIVITIES

932: OTHER AMUSEMENT AND RECREATION ACTIVITIES

I have selected the following Business types. Studying geospatial data and analysing these businesses will help us get a clear picture of how Singapore's topurism scene functions. These analysis can be used for new businesses in future to position themselves intelligently to increase revenue. It can also be used for urban planning.

```{r}
pacman::p_load(tidyverse, sf, tmap, httr, spatstat, rvest)
```

### Import Acra data

### import data and remove problematic data

```{r}
folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path, 
                        pattern = "^ACRA*.*\\.csv$", 
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)
```

## **Saving ACRA data**

save the consolidated ACRA data

```{r}
write_rds(acra_data,
          "data/rds/acra_data.rds")
```

## **Tidying ACRA data**

pad problematc postal codes

```{r}
acra_data <- acra_data %>% mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0"))
```

We will keep the first 25 columns and filter for businesses classified under SSIC codes **47 (Retail Trade)**, **551 (Short Term Accommodation Activities)**, and **932 (Other Amusement and Recreation Activities)**. Additionally, we will restrict the dataset to businesses registered between **1 January 2024** and **30 June 2025**.

```{r}

# Retail Trade (47)
biz47 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 2) == "47"
  )

# Short Term Accommodation Activities (551)
biz551 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 3) == "551"
  )

# Other Amusement and Recreation Activities (932)
biz932 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 3) == "932"
  )

```

## Fetch all unique postal

```{r}
# 551 - Short Term Accommodation Activities
postcodes_551 <- unique(biz551$postal_code)

# 47 - Retail Trade
postcodes_47 <- unique(biz47$postal_code)

# 932 - Other Amusement and Recreation Activities
postcodes_932 <- unique(biz932$postal_code)

```

## Geocoding

```{r}
library(httr)
library(dplyr)

fetch_postcodes <- function(df, url) {
  postcodes <- unique(df$postal_code)
  
  found <- data.frame()
  not_found <- data.frame(postcode = character())
  
  for (pc in postcodes) {
    query <- list(
      searchVal = pc,
      returnGeom = "Y",
      getAddrDetails = "Y",
      pageNum = "1"
    )
    
    res <- GET(url, query = query)
    json <- content(res)
    
    if (json$found != 0) {
      df_res <- as.data.frame(json$results, stringsAsFactors = FALSE)
      df_res$input_postcode <- pc
      found <- bind_rows(found, df_res)
    } else {
      not_found <- bind_rows(not_found, data.frame(postcode = pc))
    }
  }
  
  return(list(found = found, not_found = not_found))
}

```

```{r}
url <- "https://onemap.gov.sg/api/common/elastic/search"

res_47  <- fetch_postcodes(biz47, url)
res_551 <- fetch_postcodes(biz551, url)
res_932 <- fetch_postcodes(biz932, url)

```

Fetch the first 10 columns including the lat and long

```{r}
res_47$found <- res_47$found %>% select(1:10)
res_551$found <- res_551$found %>% select(1:10)
res_932$found <- res_932$found %>% select(1:10)
```

Appending the location information

```{r}
biz47 <- biz47 %>%
  left_join(res_47$found, 
            by = c("postal_code" = "POSTAL"))

biz551 <- biz551 %>%
  left_join(res_551$found, 
            by = c("postal_code" = "POSTAL"))

biz932 <- biz932 %>%
  left_join(res_932$found, 
            by = c("postal_code" = "POSTAL"))

```

Save the data

```{r}
write_rds(biz47, "data/rds/biz_47.rds")
write_rds(biz551, "data/rds/biz_551.rds")
write_rds(biz932, "data/rds/biz_932.rds")

```

Read rds

```{r}
biz47 <- readRDS("data/rds/biz_47.rds")
biz551 <- readRDS("data/rds/biz_551.rds")
biz932 <- readRDS("data/rds/biz_932.rds")

```

Convert from tibble to sf

```{r}

biz47_sf <- st_as_sf(biz47,
                     coords = c("X", "Y"),
                     crs = 3414)

biz551_sf <- st_as_sf(biz551,
                      coords = c("X", "Y"),
                      crs = 3414)

biz932_sf <- st_as_sf(biz932,
                      coords = c("X", "Y"),
                      crs = 3414)
```

## **Visualising the businesses**

```{r}
tmap_mode("view")

tm_shape(biz47_sf) +
  tm_dots(col = "blue", size = 0.1) +
tm_shape(biz551_sf) +
  tm_dots(col = "red", size = 0.1) +
tm_shape(biz932_sf) +
  tm_dots(col = "green", size = 0.1)
```

## Master Plan 2019 Subzone (No Sea) data set into R environment.

```{r}
mpsz_sf <- st_read("data/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)
```

## Convert to ppp from sf for further analysis

PPP = planar point pattern

```{r}
W <- as.owin(mpsz_sf)

biz47_ppp  <- as.ppp(st_coordinates(biz47_sf),  W = W)
biz551_ppp <- as.ppp(st_coordinates(biz551_sf), W = W)
biz932_ppp <- as.ppp(st_coordinates(biz932_sf), W = W)
```

### Kernel density across Singapore for all businesses

```{r}
# --- biz47 ---
bw_diggle_47 <- bw.diggle(biz47_ppp)
kde_biz47 <- density(
  biz47_ppp,
  sigma = bw_diggle_47,
  edge = TRUE,
  kernel = "gaussian"
)

# --- biz551 ---
bw_diggle_551 <- bw.diggle(biz551_ppp)
kde_biz551 <- density(
  biz551_ppp,
  sigma = bw_diggle_551,
  edge = TRUE,
  kernel = "gaussian"
)

# --- biz932 ---
bw_diggle_932 <- bw.diggle(biz932_ppp)
kde_biz932 <- density(
  biz932_ppp,
  sigma = bw_diggle_932,
  edge = TRUE,
  kernel = "gaussian"
)

```

### KDE for Retail (47)

```{r}
kde_biz47 <- density(biz47_ppp,
                              sigma=1000, 
                              edge=TRUE,
                              kernel="gaussian")
plot(kde_biz47)
```

### KDE for Accommodation (551)

```{r}
kde_biz551 <- density(
  biz551_ppp,
  sigma = 1000, 
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_biz551, main = "KDE - Accommodation (551)")
```

### KDE - Amusement & Recreation (932)

```{r}
kde_biz932 <- density(
  biz932_ppp,
  sigma = 1000, 
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_biz932, main = "KDE - Amusement & Recreation (932)")
```

The intensities are tiny and the window is huge. Hence we are going to analyze by region

## Extract regions from mpsz

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

remove islands

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
write_rds(mpsz_cl, 
          "data/rds/mpsz_cl.rds")
```

```{r}
# --- Retail Trade (47) ---
tm_shape(mpsz_sf) +
  tm_polygons() +
tm_shape(biz47_sf) +
  tm_dots(col = "red")
```

```{r}
# --- Accommodation (551) ---
tm_shape(mpsz_sf) +
  tm_polygons() +
tm_shape(biz551_sf) +
  tm_dots(col = "blue")
```

```{r}
# --- Amusement & Recreation (932) ---
tm_shape(mpsz_sf) +
  tm_polygons() +
tm_shape(biz932_sf) +
  tm_dots(col = "green")

```

```{r}
kde_biz47 <- density(biz47_ppp,
                              sigma=1000
                     , 
                              edge=TRUE,
                              kernel="gaussian")
plot(kde_biz47)
```
