---
title: "Take-Home_Ex01"
author: "Mohammad Farhan Ahsan"
date: "September 20, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Datasets used

Acra data

*Master Plan 2019 Subzone Boundary (No Sea)*, a polygon feature data providing information of URA 2019 Master Plan Planning Subzone boundary data.

## Business types selected:

47: Retail Trade

55: SHORT TERM ACCOMMODATION ACTIVITIES

932: OTHER AMUSEMENT AND RECREATION ACTIVITIES

I have selected the following Business types. Studying geospatial data and analysing these businesses will help us get a clear picture of how Singapore's topurism scene functions. These analysis can be used for new businesses in future to position themselves intelligently to increase revenue. It can also be used for urban planning.

```{r}
pacman::p_load(tidyverse, sf, tmap, httr, spatstat, rvest, sparr, stpp)
```

### Import Acra data

### import data and remove problematic data

```{r, eval=FALSE}
folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path, 
                        pattern = "^ACRA*.*\\.csv$", 
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)
```

## **Saving ACRA data**

save the consolidated ACRA data

```{r, eval=FALSE}
write_rds(acra_data,
          "data/rds/acra_data.rds")
```

## **Tidying ACRA data**

pad problematc postal codes

```{r, eval=FALSE}
acra_data <- acra_data %>% mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0"))
```

We will keep the first 25 columns and filter for businesses classified under SSIC codes **47 (Retail Trade)**, **551 (Short Term Accommodation Activities)**, and **932 (Other Amusement and Recreation Activities)**. Additionally, we will restrict the dataset to businesses registered between **1 January 2024** and **30 June 2025**.

```{r, eval=FALSE}

# Retail Trade (47)
biz47 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 2) == "47"
  )

# Short Term Accommodation Activities (551)
biz551 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 3) == "551"
  )

# Other Amusement and Recreation Activities (932)
biz932 <- acra_data %>%
  select(1:25) %>%
  filter(
    registration_incorporation_date >= as.Date("2024-01-01") &
    registration_incorporation_date <= as.Date("2025-06-30") &
    substr(as.character(primary_ssic_code), 1, 3) == "932"
  )

```

## Fetch all unique postal

```{r, eval=FALSE}
# 551 - Short Term Accommodation Activities
postcodes_551 <- unique(biz551$postal_code)

# 47 - Retail Trade
postcodes_47 <- unique(biz47$postal_code)

# 932 - Other Amusement and Recreation Activities
postcodes_932 <- unique(biz932$postal_code)

```

## Geocoding

```{r, eval=FALSE}
library(httr)
library(dplyr)

fetch_postcodes <- function(df, url) {
  postcodes <- unique(df$postal_code)
  
  found <- data.frame()
  not_found <- data.frame(postcode = character())
  
  for (pc in postcodes) {
    query <- list(
      searchVal = pc,
      returnGeom = "Y",
      getAddrDetails = "Y",
      pageNum = "1"
    )
    
    res <- GET(url, query = query)
    json <- content(res)
    
    if (json$found != 0) {
      df_res <- as.data.frame(json$results, stringsAsFactors = FALSE)
      df_res$input_postcode <- pc
      found <- bind_rows(found, df_res)
    } else {
      not_found <- bind_rows(not_found, data.frame(postcode = pc))
    }
  }
  
  return(list(found = found, not_found = not_found))
}

```

```{r, eval=FALSE}
url <- "https://onemap.gov.sg/api/common/elastic/search"

res_47  <- fetch_postcodes(biz47, url)
res_551 <- fetch_postcodes(biz551, url)
res_932 <- fetch_postcodes(biz932, url)

```

Fetch the first 10 columns including the lat and long

```{r, eval=FALSE}
res_47$found <- res_47$found %>% select(1:10)
res_551$found <- res_551$found %>% select(1:10)
res_932$found <- res_932$found %>% select(1:10)
```

Appending the location information

```{r, eval=FALSE}
biz47 <- biz47 %>%
  left_join(res_47$found, 
            by = c("postal_code" = "POSTAL"))

biz551 <- biz551 %>%
  left_join(res_551$found, 
            by = c("postal_code" = "POSTAL"))

biz932 <- biz932 %>%
  left_join(res_932$found, 
            by = c("postal_code" = "POSTAL"))

```

Save the data

```{r, eval=FALSE}
write_rds(biz47, "data/rds/biz_47.rds")
write_rds(biz551, "data/rds/biz_551.rds")
write_rds(biz932, "data/rds/biz_932.rds")

```

Read rds

```{r}
biz47 <- readRDS("data/rds/biz_47.rds")
biz551 <- readRDS("data/rds/biz_551.rds")
biz932 <- readRDS("data/rds/biz_932.rds")

```

Convert from tibble to sf

```{r}

biz47_sf <- st_as_sf(biz47,
                     coords = c("X", "Y"),
                     crs = 3414)

biz551_sf <- st_as_sf(biz551,
                      coords = c("X", "Y"),
                      crs = 3414)

biz932_sf <- st_as_sf(biz932,
                      coords = c("X", "Y"),
                      crs = 3414)
```

## **Visualising the businesses**

```{r}
tmap_mode("view")

tm_shape(biz47_sf) +
  tm_dots(col = "blue", size = 0.1) +
tm_shape(biz551_sf) +
  tm_dots(col = "red", size = 0.1) +
tm_shape(biz932_sf) +
  tm_dots(col = "green", size = 0.1)
```

## Master Plan 2019 Subzone (No Sea) data set into R environment.

```{r}
mpsz_sf <- st_read("data/rawdata/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)
```

```{r}
mpsz_cl <- read_rds("data/rds/mpsz_cl.rds")
```

## Convert to ppp from sf for further analysis

PPP = planar point pattern

```{r}
W <- as.owin(mpsz_sf)

biz47_ppp  <- as.ppp(st_coordinates(biz47_sf),  W = W)
biz551_ppp <- as.ppp(st_coordinates(biz551_sf), W = W)
biz932_ppp <- as.ppp(st_coordinates(biz932_sf), W = W)
```

### Kernel density across Singapore for all businesses

```{r}
# --- biz47 ---
bw_diggle_47 <- bw.diggle(biz47_ppp)
kde_biz47 <- density(
  biz47_ppp,
  sigma = bw_diggle_47,
  edge = TRUE,
  kernel = "gaussian"
)

# --- biz551 ---
bw_diggle_551 <- bw.diggle(biz551_ppp)
kde_biz551 <- density(
  biz551_ppp,
  sigma = bw_diggle_551,
  edge = TRUE,
  kernel = "gaussian"
)

# --- biz932 ---
bw_diggle_932 <- bw.diggle(biz932_ppp)
kde_biz932 <- density(
  biz932_ppp,
  sigma = bw_diggle_932,
  edge = TRUE,
  kernel = "gaussian"
)

```

### KDE for Retail (47)

```{r}
kde_biz47 <- density(
  biz47_ppp,
  sigma=1000, 
  edge=TRUE,
  kernel="gaussian"
)
plot(kde_biz47)
```

### KDE for Accommodation (551)

```{r}
kde_biz551 <- density(
  biz551_ppp,
  sigma = 1000, 
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_biz551, main = "KDE - Accommodation (551)")
```

### KDE - Amusement & Recreation (932)

```{r}
kde_biz932 <- density(
  biz932_ppp,
  sigma = 1000, 
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_biz932, main = "KDE - Amusement & Recreation (932)")
```

The intensities are tiny and the window is huge. Hence we are going to analyze by region

## Spatio temporal KDE

by month for biz 47

```{r}


# 1) Build ordered month index (numeric) + a nice label to use in plots
biz47_sf <- biz47_sf %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),         # Date at 1st of month
    YearMonthLab  = format(YearMonthDate, "%Y-%m")               # "YYYY-MM" label
  ) %>%
  mutate(
    Month_idx = as.integer(factor(YearMonthDate,
                                  levels = sort(unique(YearMonthDate))))
  )
# Month_idx = 1,2,3,... across the whole 1.5 years

# 2) Keep only the temporal mark + geometry (sf keeps geometry)
biz47_month <- biz47_sf %>%
  select(Month_idx)

# 3) Convert to ppp (bounding box window for now)
biz47_month_ppp <- as.ppp(biz47_month)
summary(biz47_month_ppp)
any(duplicated(biz47_month_ppp))

# 4) Clip to Singapore boundary (owin from your cleaned MPSZ)
sg_owin <- as.owin(mpsz_cl)
biz47_month_owin <- biz47_month_ppp[sg_owin]
summary(biz47_month_owin)


plot(biz47_month_owin)
```

```{r}
# 5) Spatio-temporal KDE (auto bandwidths; or set h=..., lambda=...)
biz47_stkde <- spattemp.density(biz47_month_owin)
summary(biz47_stkde)



```

```{r}
# 6) Plot all months sequentially, full size, with proper YYYY-MM labels
#    Build a lookup from index -> label
idx_levels <- sort(unique(biz47_sf$Month_idx))
lab_lookup <- biz47_sf %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)

for (i in idx_levels) {
  plot(biz47_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz47 (", lab_lookup$YearMonthLab[lab_lookup$Month_idx == i], ")", sep = ""))
}



```

```{r}
# --- biz551 ----------------------------------------------------
biz551_sf <- biz551_sf %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),
    YearMonthLab  = format(YearMonthDate, "%Y-%m"),
    Month_idx = as.integer(factor(YearMonthDate,
                                  levels = sort(unique(YearMonthDate))))
  )

biz551_month <- biz551_sf %>%
  select(Month_idx)

biz551_month_ppp <- as.ppp(biz551_month)
summary(biz551_month_ppp)
any(duplicated(biz551_month_ppp))

biz551_month_owin <- biz551_month_ppp[sg_owin]
summary(biz551_month_owin)

biz551_stkde <- spattemp.density(biz551_month_owin)
summary(biz551_stkde)

# Plot all months with labels
idx_levels_551 <- sort(unique(biz551_sf$Month_idx))
lab_lookup_551 <- biz551_sf %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)

for (i in idx_levels_551) {
  plot(biz551_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz551 (", 
                    lab_lookup_551$YearMonthLab[lab_lookup_551$Month_idx == i], 
                    ")", sep = ""))
}

```

```{r}
# --- biz932 ----------------------------------------------------
biz932_sf <- biz932_sf %>%
  mutate(
    YearMonthDate = floor_date(uen_issue_date, "month"),
    YearMonthLab  = format(YearMonthDate, "%Y-%m"),
    Month_idx = as.integer(factor(YearMonthDate,
                                  levels = sort(unique(YearMonthDate))))
  )

biz932_month <- biz932_sf %>%
  select(Month_idx)

biz932_month_ppp <- as.ppp(biz932_month)
summary(biz932_month_ppp)
any(duplicated(biz932_month_ppp))

biz932_month_owin <- biz932_month_ppp[sg_owin]
summary(biz932_month_owin)

biz932_stkde <- spattemp.density(biz932_month_owin)
summary(biz932_stkde)

# Plot all months with labels
idx_levels_932 <- sort(unique(biz932_sf$Month_idx))
lab_lookup_932 <- biz932_sf %>%
  distinct(Month_idx, YearMonthLab) %>%
  arrange(Month_idx)

for (i in idx_levels_932) {
  plot(biz932_stkde, i,
       override.par = FALSE,
       fix.range    = TRUE,
       main = paste("ST-KDE – biz932 (", 
                    lab_lookup_932$YearMonthLab[lab_lookup_932$Month_idx == i], 
                    ")", sep = ""))
}
```

## 2nd order spatial analytics

### Retail

Check for competition

```{r}
set.seed(1234)
```

```{r}
G_biz47  = Gest(biz47_ppp, correction = "border")
plot(G_biz47 , xlim=c(0,500))
```

```{r, eval=FALSE}
G_biz47_csr <- envelope(biz47_ppp, Gest, nsim = 999)

plot(G_biz47_csr, 
     main = "Monte Carlo CSR test for Retail Businesses (G-function)",
     xlab = "Distance (m)", ylab = "G(r)")

```

### Accomodation

```{r}
set.seed(1234)
```

```{r}
# Empirical G with border correction
G_biz551 <- Gest(biz551_ppp, correction = "border")
```

```{r}
# Quick look (limit to 0–500 m like your Retail plot)
plot(G_biz551, xlim = c(0, 500),
     main = "G-function – Accommodation (551)",
     xlab = "Distance r (m)", ylab = "G(r)")
```

```{r}
G_biz551_env <- envelope(
  biz551_ppp,
  fun       = Gest,
  nsim      = 999,          # use 999 for final report
)


```

```{r}
plot(G_biz551_env,
     main = "Monte Carlo CSR test – G(r) for Accommodation (551)",
     xlab = "Distance r (m)", ylab = "G(r)")
```

### Attractions

```{r}
set.seed(1234)


```

```{r}
F_biz932 <- Fest(biz932_ppp, correction = "border")


```

```{r}
plot(F_biz932, xlim = c(0, 500),
     main = "F-function – Amusement & Recreation (932)",
     xlab = "Distance r (m)", ylab = "F(r)")
```

```{r}
F_biz932_env <- envelope(
  biz932_ppp,
  fun       = Fest,
  nsim      = 999,          # use 999 for final report
)

```

```{r}
plot(F_biz932_env,
     main = "Monte Carlo CSR test – F(r) for Amusement & Recreation (932)",
     xlab = "Distance r (m)", ylab = "F(r)")
```

### Summary

From the G-function analysis, we observe that retail businesses (SSIC code 47) exhibit clustering behavior, as indicated by the empirical G-function lying above the CSR envelope. This suggests that retail businesses tend to be located closer to each other than would be expected under a random distribution. In contrast, the accommodation businesses (SSIC code 551) show a more random spatial distribution, with the empirical G-function closely following the CSR envelope. This indicates that accommodation providers are more evenly spread out across the area, likely to cater to a wider range of locations for tourists. The F-function analysis for amusement and recreation businesses (SSIC code 932) reveals a tendency towards regularity, as the empirical F-function lies below the CSR envelope. This suggests that these businesses are more evenly spaced than would be expected under a random distribution, possibly to avoid direct competition and to cover different market segments. Overall, the spatial patterns of these business types reflect their differing operational strategies and market demands. Retail businesses cluster to attract foot traffic, accommodation providers distribute evenly to maximize accessibility, and amusement and recreation businesses space themselves out to minimize competition.
